Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-03-26 Dail_Mad_ext.tex
Some trimming

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a
patch of early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located survey plots. The time
frame of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a
one month long breeding season, which could be surveyed once per
year for $T$ years. If the a site is surveyed on $J$ occasions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e.
$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. If secondary sampling periods were
used, an extra dimension is added so that we have $X_{ijt}:
j=1,\hdots,J$. Typically, we know $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss2},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process
simply describes the the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

To make the connection between conventional state-space models
and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
Eq.~\ref{eq:ss2b}, with an expression derived from
Eq.~\ref{eq:Nt}. This requires summing over all possible values
of $S$ and $G$, which is accomplished using the discrete convolution:
\begin{equation}
[N_{it}|N_{it-1}, \omega, \gamma] =
  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
\times
\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
\end{equation}
where here we assumed no dependence of $\gamma$ on $N_{it-1}$.

In the absence of movement, $\omega$ is exactly the probability of surviving
from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
(under the geometric growth model). This ability to directly estimate
demographic parameters from simple count data is one of the DM
model's most appearling features. However,
in the more common scenario when immigration and emigration
occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
rates. Rather, $\omega$ is the probability of surviving and not
emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
i.e. the recruitment rate. In such cases, we propose
replacing Eq.~\ref{eq:Nt} with standard population models.

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

The entire model can now be written:
\begin{gather}
  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\  
G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
  N_{it} = S_{it}+G_{it} \nonumber \\
  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
  \label{eq:hm}
\end{gather}

\section{Statistical Inference}

\subsection{Maximum likelihood}

Maximum likelihood estimation of parameters in models with
discrete random effects requires removing them by
summation. For Markovian models, this can be done recursively,
making the likelihood tractable. A general expression of the likelihood
can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $N_{max}$, which
%%% Can we rename the upper bound?  We're also using K
%%% for equilibrium abundance.
should be high enough such that the MLEs are not affected by increasing
it further. This likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

@@

Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data. [EXPLAIN POSTERIORS AND PRIORS]
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. Although MCMC can be slow, our experience is that it
may actually converge faster than the amount of time required to
maximize the likelihood shown in Eq.\ref{eq:like}, especially
when the upper bound of the summation must be high, e.g. $N_{max}>200$.
Furthermore, MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing
users to specify the model using a simple symbolic descriptions.

@@

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. Although demographic and environmental stochasticity
in $r$ could be modeled as functions of covariates or random effects, a
simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

The autoregressive, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  This model is close to
the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
of $\omega$ and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The autoregressive, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the autoregressive +  immigration model as: (14)
\end{comment}

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth
terms. For example, the recruitment term of the constant DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

@@

One way to model environmental stochasticity (variation in population growth
rate due to stochastic environmental conditions) is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature and $K$ a function of rainfall:
\begin{gather}
r_{it} = \exp(r_{0} + r_{temp}T_{it}) \nonumber \\
K_{it} = \exp(K_{0} + K_{rain}R_{it})  \\
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
\label{eq:weather}
\end{gather}
where $T_{it}$ is site and year specific temperature, $R_{it}$ is site and year specific rainfall,
and the errors for the coefficients $r_{0}$, $r_{temp}$, $K_{0}$, and $K_{rain}$ are 
normally distributed.

@@

State-space models allow for inference about future population size,
a critical component of population viability analysis as well
as conservation planning. Let $N^+_{t}$ denote population size at some
future point in time and $[N^+_t]$ denote the probability
distribution for this random variable. If we can specify this
distribution, we can make statements about the
probability that $N^+_t = k$ where $k$ is any possible
population size including zero.

The classical approach to computing $[N^+_t]$ is based upon
Empirical Bayes methods [CITE]. 
%Note that neither Empiriis not actually a
%Bayesian method, it is used to estim.
%We take the parameter estimates and plug them back in.... For
%example to obtain the posterior distribution for $[N_{i1}]$
%\begin{equation}
%[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
%\hat{p}]}{\sum_{N_{i1}=0}^K
%[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
%  \label{eq:eb1}
%\end{equation}
%Once we have estimated the posterior distribution for the first
%time period, we can then proceed to estimate the posterior
%distribution for subsequent time periods using recursion.
%\begin{equation}
%  [\hat{N}_{it}] =
%  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
%\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
%\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
%  \label{eq:eb2}
%\end{equation}
%
%Note that when $X_{it}$ is not observed, including for future
%time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
%uniform such that it has no influence on the posterior. In other words,
%the posterior for years with no observed data is determined entirely
%by the estimated posterior from the previous year and the
%population dynamics parameters.
%
If covariates of the dynamics parameters are present, the problem
becomes much harder because, to fully account for uncertainty,
we need a model for the future values of the covariates as well. When
the covariates are continuous, this requires integrating over their
possible values, which becomes very challenging
computationally. Indeed, simulation methods such as MCMC are the
only viable options when numerous covariates are present, which will
often be the case in many ecological studies.

Relative to the computational challenges associated with the
empirical Bayes methods, the Bayesian approach is much more easy to
implement, especially since MCMC will be generally be relied upon in the
first place. Furthermore, because population size at any future point
in time is regarded as an unknown parameter like any other, computing the
posterior distribution of $N^+_t$ requires no additional steps
in MCMC algorithm, other than formally stating that the count data are
``missing'' in future years. An example of this is provided in the appendix.

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each
species, and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant, autoregressive, Ricker-logistic, Gompertz-logistic, autoregressive +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350 for golden-winged warblers.

>>>>

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
Typically, we know $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss2},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

%To make the connection between conventional state-space models
%and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
%Eq.~\ref{eq:ss2b}, with an expression derived from
%Eq.~\ref{eq:Nt}. This requires summing over all possible values
%of $S$ and $G$, which is accomplished using the discrete convolution:
%\begin{equation}
%[N_{it}|N_{it-1}, \omega, \gamma] =
%  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
%\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
%\times
%\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
%\end{equation}
%where here we assumed no dependence of $\gamma$ on $N_{it-1}$.
%
%In the absence of movement, $\omega$ is exactly the probability of surviving
%from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
%(under the geometric growth model). This ability to directly estimate
%demographic parameters from simple count data is one of the DM
%model's most appearling features. However,
%in the more common scenario when immigration and emigration
%occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
%rates. Rather, $\omega$ is the probability of surviving and not
%emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
%i.e. the recruitment rate. In such cases, we propose
%replacing Eq.~\ref{eq:Nt} with standard population models.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

%The entire model can now be written:
%\begin{gather}
%  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
%S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\  
%G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
%  N_{it} = S_{it}+G_{it} \nonumber \\
%  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
%  \label{eq:hm}
%\end{gather}

%\subsection{Statistical Inference}

%\subsubsection{Maximum likelihood}

%Maximum likelihood estimation of parameters in models with
%discrete random effects requires removing them by
%summation. For Markovian models, this can be done recursively,
%making the likelihood tractable \citep{dail_madsen:2011}. 
%A general expression of the likelihood can be written:
%\begin{multline}
%  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
%  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
%  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%%  \prod_{t=T}^2 \left\{
%  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
%  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
%  \right\} \right\}
%  \label{eq:like}
%\end{multline}
%where $\infty$ is replaced by a finite upper bound $N_{max}$, which
%%% Can we rename the upper bound?  We're also using K
%%% for equilibrium abundance.
%should be high enough such that the MLEs are not affected by increasing
%it further. 
%The likelihood can be maximized numerically using the
%\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
%are given in the Appendix.

@@

\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
likelihood estimation. The likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

Writing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as %\textbf{WinBUGS} and
\textbf{JAGS} \citep{plummer:2003} overcome the technical 
difficulties by allowing users to specify the model using a simple symbolic 
descriptions. Examples are given in the Appendix.

@@

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. Although demographic and environmental stochasticity
in $r$ could be modeled as functions of covariates or random effects, a
simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  This model is close to
the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
of $\omega$ and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth
terms. For example, the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

@@

One way to model environmental stochasticity (variation in population growth
rate due to stochastic environmental conditions) is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature ($T$) and $K$ a function of rainfall ($R$):
\begin{gather}
\log(r_{it}) = \beta^{r}_{0} + \beta^{r}_{1}T_{it} \nonumber \\
\log(K_{it}) = \beta^{K}_{0} + \beta^{K}_{1}R_{it}  
%N_{it} \sim
%\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
\label{eq:weather}
\end{gather}

@@

State-space models allow for inference about future population size,
a critical component of population viability analysis (PVA) as well
as conservation planning. Let $N^+_{t}$ denote population size at some
future point in time and $[N^+_t]$ denote the probability
distribution for this random variable. If we can specify this
distribution, we can make statements about the
probability that $N^+_t = k$ where $k$ is any possible
population size including zero.

%The classical approach to computing $[N^+_t]$ is based upon
%Empirical Bayes methods [CITE]. 
%Note that neither Empiriis not actually a
%Bayesian method, it is used to estim.
%We take the parameter estimates and plug them back in.... For
%example to obtain the posterior distribution for $[N_{i1}]$
%\begin{equation}
%[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
%\hat{p}]}{\sum_{N_{i1}=0}^K
%[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
%  \label{eq:eb1}
%\end{equation}
%Once we have estimated the posterior distribution for the first
%time period, we can then proceed to estimate the posterior
%distribution for subsequent time periods using recursion.
%\begin{equation}
%  [\hat{N}_{it}] =
%  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
%\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
%\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
%  \label{eq:eb2}
%\end{equation}
%
%Note that when $X_{it}$ is not observed, including for future
%time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
%uniform such that it has no influence on the posterior. In other words,
%the posterior for years with no observed data is determined entirely
%by the estimated posterior from the previous year and the
%population dynamics parameters.
%
%If covariates of the dynamics parameters are present, the problem
%becomes much harder because, to fully account for uncertainty,
%we need a model for the future values of the covariates as well. When
%the covariates are continuous, this requires integrating over their
%possible values, which becomes very challenging
%computationally. Indeed, simulation methods such as MCMC are the
%only viable options when numerous covariates are present, which will
%often be the case in many ecological studies.

%% COULD INCLUDE THIS PARAGRAPH, OR NOT.
%A standard approach to PVA uses previously derived estimates of current 
%population size, vital rate or population growth parameters, and environmental
%stochasticity to stochastically project the population forward and calculate probabilities
%of extinction or other measures of risk.  Estimates from DM models and the extensions
%given here can be used in this context.  Uncertainty in extinction or quasi-extinction probabilities
%can be estimated by either drawing values for parameters from their posterior distributions
%(if estimated using MCMC) or randomly drawing values for parameters using their means and standard
%errors (if estimated using maximum likelihood), and running a number of simulations with each
%set of parameter values \citep{white:2000,hostetler_etal:2012}.

%Relative to the computational challenges associated with the
%empirical Bayes methods, 
A fully Bayesian approach to PVA is straightforward to
implement, especially since MCMC will be generally be relied upon in the
first place. Because population size at any future point %Furthermore, b
in time is regarded as an unknown parameter like any other, computing the
posterior distribution of $N^+_t$ requires no additional steps
in MCMC algorithm, other than formally stating that the count data are
``missing'' in future years. An example of this is provided in the appendix.

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each
species, and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350 for golden-winged warblers.
